Union and intersection contracts: theory vs practice

### 2021-02-18 paper.tex
WIP

\documentclass{article}
\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}

@@

\section{First section}
Some text

\section{Second section}
Some text

\section{Third section}
Some text

\section*{Conclusion}
Conclusion

>>>>

\documentclass{article}
\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}

@@

\section{Background on union types}

\subsection{Nickel design space}

\color{red}Goals of the section

Introduce Nickel, explain the choice of gradual
typing and main design orientations, such as practice-oriented, lightweight,
etc.\vspace{0.5cm}\color{black}

This paper is centered around the design of the Nickel language. Nickel is a
configuration programming language, meaning that a program eventually evaluates
to a value that can be directly serialized to JSON, YAML or alike. Nickel aims at
being lightweight, easy to write and easy to embed in a larger application. The
focus is first and foremost practical: it originated from the language of the Nix
package manager, used to describe the biggest set of software packages ever
built [ ], and targets build systems and cloud deployment tools (Kubernetes,
Terraform or NixOps) as well. While there's no hesitation in incorporating recent
research when it serves a purpose, this is not a research language, as the
design decisions are guided by the industrial use cases.

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML.  Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. There's a dilemma with respect to typing: if
your program is only run once on fixed inputs, as are such programmable
configurations, any type error will be reported at run-time anyway. Why bother
with a static type system? On the other hand, more and more systems rely on
complex configurations, like cloud infrastructure, leading the corresponding
configurations to become increasingly complex, to the point where static types
are beneficial. For reusable code — that is, library functions — static types
brings all the usual, widely acknowledged benefits. This dilemma is naturally
solved by an approach that can reunite static typing and dynamic typing: gradual
typing [references]. Unlike language like TypeScript, we don't seek to provide a
complex type system that tries hard to type most of the idioms arising naturally
in dynamically typed code. Rather, we provide a reasonably solid base type
system with good inference properties, that should make writing typed generic
function operating on data easy. In fact, for more complex data validation,
Nickel uses contracts rather static typing.

\subsection{Unions types}
\color{red}Goals of the section

Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}\color{black}

In programming, unions are fundamental and ubiquitous building blocks of data.
While products represent conjunctions, which are several different pieces of
data stored all together in one value, unions represent disjunctions, which are
several exclusive alternatives stored in one value. When nested together, unions
and intersections are able to describe a large class of data shapes [ref,
examples?].

\begin{verbatim}
EXAMPLES of union in configuration specification
\end{verbatim}

In statically typed languages, and more specifically in functional programming
languages, unions are usually embodied by algebraic data types [references],
also called \textbf{tagged unions}. Tagged means that there is an additional
field, or constructor, that allows to discriminate between each alternative,
usually via pattern matching. Well-known programming languages with first-class
tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm, Swift and an
even greater number of related and derived languages.

\begin{verbatim}
EXAMPLES
\end{verbatim}

\subsection{Tagged versus untagged}
\color{red}Goals of the section

Interpreted languages already need to tag their values, hence bare union are more natural.
It also corresponds better to prior usage in untyped code\vspace{0.5cm}\color{black}

In this statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors, destructors and a
well-defined semantics.  However, in a gradually typed language like Nickel, the
question of the representation of tagged unions as untyped values arises.

Unfortunately, it seems that tagged unions do not have a canonical counterpart
as serializable untyped data. An intuitive representation is as a record with a
tag together with a value which shape depends on the tag:

\begin{lstlisting}
{ tag: <A, B, C>, value: Dyn }
\end{lstlisting}

Another way of seeing it is a dependent $\Sigma$-type which first component is a
tag and second component is the data type corresponding to this tag:
\[
    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
\]

Alas, choosing this representation as a default is not fully satisfying. As any
representation, it does not correspond precisely to, say, a JSON type. In
consequence, a lot of illegal combinations are structurally valid untyped (JSON)
values. More importantly, this does not match with the configuration schemas one
can find in the wild favors simple, untagged union [example of JSON
schemas/protobuf/whatever with unions].

In dynamically typed languages, which are interpreted for the vast majority
[ref/examples?], it is more natural to directly stores different kind of data in
one variable without further ceremony. The runtime system needs to attach type
tags to values anyway, which usually are accessible via functions like the
JavaScript's \verb+typeof+. Using so-called untagged unions has already been an
idiom in dynamically typed language for a long time [ref/examples], and being
able to type and handle these unions an important aspect of the migration to a
gradual type system.

Most of gradually typed languages either existed originally as dynamically typed
language or aim at backward compatibility with a dynamically typed base
language. This is the reason they tend to favor supporting union types, the
typed model of the practice of simply storing different kind of data in one
variable [example of gradually typed language with untagged union].

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. But are we? As a configuration language,
Nickel does not evolve in the abstract. It aims at best interoperability with
JSON and alike. It is designed to generate data for other softwares to consume,
and tries as much as possible to be able to encode data schemas at the language
level. If these schemas use untagged union, then this is a strong incentive to
support them at the type level, or at least, at the contract level.

\subsection{Union types}

\subsection{Tagged as untagged}
\color{red}Goals of the section

Untagged unions, modulo a slighty smarter typechecker, can encode usual type-safe
tagged unions. They let the user free of the representation though, and encode
more stuff, that's why they are appealing.\vspace{0.5cm}\color{black}

Untagged union, together with the other existing features of the type system,
are sufficient to implement the encoding mentioned in [cite previous section]:

Encoding as: use singleton contract for tags, and then union.
\begin{lstlisting}
type sum = A of Num | B of Str
\end{lstlisting}

\begin{lstlisting}
{ tag : <A>, val : Num} | { tag : <B>, val : Str}
\end{lstlisting}

As is, the static type system able to refine the type of the value though,
making the following example rejected:

\begin{lstlisting}
switch x.tag {
  A => x.val + 1,
  B => then strings.fromNum x.val
}
\end{lstlisting}

But TypeScript is capable of typing this kind of things [ ], and one can imagine
doing the same in Nickel. Equipped with tagged union and a typechecker smart
enough, classic tagged unions can be simulated. But the representation is not
imposed by the language. Untagged unions empower the programmer to represent and
handle a wider variety of unions appearing in the wild. Thus, adding union types
(and their natural and unavoidable [ref??] counterpart, intersections) to Nickel
seems very appealing.

\subsection{Union and intersections}
\color{red}Goals of the section

Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}\color{black}
